The difference operation between templates of binary cellular automata

### 2015-11-19 article/article.tex
minor fix

According to Boccara and Fukś (\cite{boccara2002}), a CA is conservative when each of its local function $f$, applied in every neighborhood $(\alpha_0,\alpha_1, \dots, \alpha_{n-1})$ satisfy the conditions described in Eq. \ref{eq:conservativeCA}.
\begin{equation}
\begin{split}
f(\alpha_0,\alpha_1, \dots,\alpha_{n-1}) = \alpha_0 + (\sum_{i=0}^{n-2}f(0_0,0_1, \dots,0_i,\alpha_1,\alpha_2, \dots,\alpha_{n-1}) \\- f(0_0,0_1, \dots,0_i,\alpha_0,\alpha_1, \dots,\alpha_{n-i-1}))
\label{eq:conservativeCA}
\end{split}
\end{equation}

@@

@@

  \begin{figure}[h!]
    \centering
    \includegraphics[width=.5\textwidth]{fig_ruleIcon60.png}
    \caption{Transitions table of the rule 60 of the elementary space.}
    \label{fig:table60}
  \end{figure}

  \begin{figure}[h!]
    \centering
    \includegraphics[width=.5\textwidth]{fig_ruleIcon102.png}
    \caption{Transitions table of the rule 102 from the elementary space, obtained through transformation by reflection applied to the transitions table of rule 60.}
    \label{fig:table102}
  \end{figure}

% Tanto a regra 60, como as regras 102, 153 e 195 pertencem a mesma classe de equivalência dinâmica. Uma forma interessante de entender o porquê é analisando a evolução espaço-temporal dessas regras, ilustrada pela Figura \ref{fig:dynamicEquivalecy}.

%   \begin{figure}[h!]
%     \centering
%     \def\svgscale{0.45}
%     \import{../img/}{fig_ruleEquivalence60.pdf_tex}
%     \caption{Evolução espaço-temporal das regras pertencente à mesma classe dinâmica da regra 60.}
%     \label{fig:dynamicEquivalecy}
%   \end{figure}

@@

\section{Templates}
\label{sec:templates}
Template is a generalization of the state transition tables of CAs through variables. A template can represents a rule set, which in turn can presents particular property. The templates were created and implemented by De Oliveira and Verardo (2014) and are currently available at the library \textit{open source CATemplates} \cite{CATemplates} on GitHub.

@@

As an example, the template of the elementary space $T_1 = (1,1,1,1,1-x_1,x_2,x_1,0)$ represents all the rules which have at position 0 (always from right to left) state 0, at positions 4, 5, 6 and 7 the state 1, at positions 1 and 2 any state in the interval $ [0, k-1] $ and in position 3 the complementary state to the value of position 1. Therefore the template $T_1$ represents the set of elementary cellular automata $\{(1,1,1,1,1,0,0,0),(1,1,1,1,0,0,1,0),(1,1,1,1,1,1,0,0),(1,1,1,1,0,1,1,0)\}$, or in its decimal form $\{248,242,252,246\}$. The process of finding all the rules represented by a template is called \textit{expansion}.

There implemented in the library \textit{CATemplates} \cite{CATemplates} several generators algorithms of templates representing rules with certain property. The template $T_{comp} = (1 - x_0, 1 - x_4, 1 - x_2, x_4, 1 - x_1, x_2, x_1, x_0)$ of the elementary CAs, for example, when expanded generates only rules that have a maximum internal symmetry by composition. But the template $T_{inv} = (1 - x_0, 1 - x_1, 1 - x_2, 1 - x_3, x_3, x_2, x_1, x_0)$ of the elementary CAs expands only the rules with color blind property.

Besides the possibility of expansion of templates, it has been developed by De Oliveira e Verardo (2014) an algorithm to generate templates that represent the intersection between two sub-spaces of rules represented by templates. To illustrate, consider the search for a rule set that are both color blind and have maximum internal symmetry by composition. To find this rule set is necessary to perform two steps: first one must match the two templates that represent the desired properties, thus forming an equation system; then to solve the equations, relationships are generated between the variables that, when applied to templates received as input will result in the template representing the intersection.

@@

For example, we will continue the difference operation between $T_{inv}$ and $T_{con}$; to this end, consider the template $T_i = (1, 1 - x_1, 1 - x_2, 1 - x_1 - x_2, x_1 + x_2, x_2, x_1, 0)$ the intersection between them, for $k=2$.
The first step of the difference operation occurs normally and generate the templates $\{(x_7, x_6, x_5, x_4, x_3, x_2, x_1, 1),(x_7, x_6, x_5, x_1 + x_2, x_3, x_2, x_1, x_0)\}$.
But it is trivial to realize that any expansion of $T_i$ having the set of substitutions $\{x_1 = 1, x_2 = 1\}$ will cause the position $3$ and $4$ of the template display values that do not belong to the interval $[0,k-1]$.
Hence all templates that have $\{x_1 = 1, x_2 = 1\}$ are complementary to the template $T_i$. So it will generate the template $(x_7, x_6, x_5, x_4, x_3, 1, 1, x_0)$, which is the exception template of $T_i$.

Therefore every rule represented  by the template $T_{minuend}$ -- $T_{inv}$ in the given example -- and which is also represented by the exception template from the the intersection of $T_{minuend}$ with $T_{subtrahend}$ -- $T_i$ in this example - - should be represented by at least one of the templates resulting from the difference operation. For this, the algorithm that finds the difference between templates considers all exception templates found, intersects them with the $T_{minuend}$ and adds them to the set of templates obtained by the first steps of the difference operation. Thus, for the example presented above, the set of difference templates resultant is represented by Eq. \ref{eq:differenceR}:\begin{equation}
\begin{split}
\{(0, 1 - x_1, 1 - x_2, 1 - x_3, x_3, x_2, x_1, 1), \\
(1 - x_0, 1 - x_1, 1 - x_2, x_1 + x_2, 1 - x_1 - x_2, x_2, x_1, x_0), \\
(1 - x_0, 0, 0, 1 - x_3, x_3, 1, 1, x_0)\}
\label{eq:differenceR}
\end{split}
\end{equation}

@@

\section*{Acknowledgements}
\label{sec:agrdecimentos}
The authors thank Mackenzie Research Fund (MACKPESQUISA), the FAPESP, and federal agencies CAPES and CNPq for different forms of support received during the development of this work.

%
% ---- Bibliography ----
%
\begin{thebibliography}{}
\providecommand{\abntrefinfo}[3]{}
\providecommand{\abntbstabout}[1]{}
\abntbstabout{1.52 }

\bibitem[Betel, {DE OLIVEIRA} e Flocchini 2013]{Betel2013}
\abntrefinfo{Betel, {DE OLIVEIRA} e Flocchini}{BETEL; {DE OLIVEIRA};
  FLOCCHINI}{2013}
{BETEL, H.; {DE OLIVEIRA}, P. P.~B.; FLOCCHINI, P. Solving the parity problem
  in one-dimensional cellular automata.
\emph{Natural Computing}, v.~12, n.~3, p. 323--337, 2013.}

\bibitem[Boccara e {FUK\'{S}} 2002]{boccara2002}
\abntrefinfo{Boccara e {FUK\'{S}}}{BOCCARA; {FUK\'{S}}}{2002}
{BOCCARA, N.; {FUK\'{S}}, H. Number-conserving cellular automaton rules.
\emph{Fundamenta Informaticae}, IOS Press, v.~52, n.~1-3, p. 1--13, 2002.}

\bibitem[{DE OLIVEIRA} 2014]{deOliveira2014density}
\abntrefinfo{{DE OLIVEIRA}}{{DE OLIVEIRA}}{2014}
{{DE OLIVEIRA}, P. P.~B. On density determination with cellular automata:
  Results, constructions and directions.
\emph{Journal of Cellular Automata}, v.~9, n.~5-6, p. 357--385, 2014.}

\bibitem[{DE OLIVEIRA} e Verardo 2014]{deOliveira2014}
\abntrefinfo{{DE OLIVEIRA} e Verardo}{{DE OLIVEIRA}; VERARDO}{2014a}
{{DE OLIVEIRA}, P. P.~B.; VERARDO, M. Representing families of cellular
  automata rules.
\emph{The Mathematica Journal}, v.~16, n.~8, 2014.
Dispon{\'\i}vel em: $<$dx\-.doi\-.org\-/doi:10\-.3888\-/tmj\-.16-8$>$.}

\bibitem[{DE OLIVEIRA} e Verardo 2014]{deOliveira2014b}
\abntrefinfo{{DE OLIVEIRA} e Verardo}{{DE OLIVEIRA}; VERARDO}{2014b}
{{DE OLIVEIRA}, P. P.~B.; VERARDO, M. Template based representation of cellular
  automata rules. In:  ISOKAWA, T.; IMAI, K.; MATSIU, N.; PEPER, F.; UMEO, H.
  (Ed.). \emph{20th International Workshop on Cellular Automata and Discrete
  Complex Systems}. Himeji, Japão, Julho 7-9: [s.n.], 2014. p. 199--204.}

\bibitem[Salo e T{\"o}rm{\"a} 2013]{salo2013color}
\abntrefinfo{Salo e T{\"o}rm{\"a}}{SALO; T{\"O}RM{\"A}}{2013}
{SALO, V.; T{\"O}RM{\"A}, I. Color blind cellular automata. In:  \emph{Cellular
  Automata and Discrete Complex Systems}. [S.l.]: Springer, 2013. p. 139--154.}

\bibitem[Verardo e {DE OLIVEIRA} 2015]{CATemplates}
\abntrefinfo{Verardo e {DE OLIVEIRA}}{VERARDO; {DE OLIVEIRA}}{2015}
{VERARDO, M.; {DE OLIVEIRA}, P. P.~B. \emph{CATemplates}.
[S.l.], 2015. Dispon{\'\i}vel em:
  $<$https:\-/\-/github\-.com\-/mverardo\-/CATemplates$>$.}

\bibitem[Wolfram 1994]{wolfram1994cellular}
\abntrefinfo{Wolfram}{WOLFRAM}{1994}
{WOLFRAM, S. \emph{Cellular automata and complexity: collected papers}. [S.l.]:
  Addison-Wesley Reading, 1994.}

\bibitem[Wolfram 2002]{wolfram2002}
WOLFRAM 2002
{WOLFRAM, S. \emph{A new kind of science}. [S.l.]: Wolfram media Champaign,
  2002.}

\bibitem[Wolz e {DE OLIVEIRA} 2008]{wolz2008very}
\abntrefinfo{Wolz e {DE OLIVEIRA}}{WOLZ; {DE OLIVEIRA}}{2008}
{WOLZ, D.; {DE OLIVEIRA}, P. P.~B. Very effective evolutionary techniques for
  searching cellular automata rule spaces.
\emph{J. Cellular Automata}, v.~3, n.~4, p. 289--312, 2008.}

\end{thebibliography}
\end{document}

>>>>

According to Boccara and Fukś \cite{boccara2002}, a CA is conservative when each of its local function $f$, applied in every neighborhood $(\alpha_0,\alpha_1, \dots, \alpha_{n-1})$ satisfy the conditions described in Eq. \ref{eq:conservativeCA}.
\begin{equation}
\begin{split}
f(\alpha_0,\alpha_1, \dots,\alpha_{n-1}) = \alpha_0 + (\sum_{i=0}^{n-2}f(0_0,0_1, \dots,0_i,\alpha_1,\alpha_2, \dots,\alpha_{n-1}) \\- f(0_0,0_1, \dots,0_i,\alpha_0,\alpha_1, \dots,\alpha_{n-i-1}))
\label{eq:conservativeCA}
\end{split}
\end{equation}

@@

  \begin{figure}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{fig_ruleIcon60.png}
    \caption{Transitions table of the rule 60 of the elementary space.}
    \label{fig:table60}
  \end{figure}
  \begin{figure}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{fig_ruleIcon102.png}
    \caption{Transitions table of the rule 102 from the elementary space, obtained through transformation by reflection applied to the transitions table of rule 60.}
    \label{fig:table102}
  \end{figure}

@@

@@

\section{Templates}
\label{sec:templates}
Template is a generalization of the state transition tables of CAs through variables. A template can represents a rule set, which in turn can presents particular property. The templates were created and implemented by De Oliveira and Verardo \cite{deOliveira2014} and are currently available at the library \textit{open source CATemplates} \cite{CATemplates} on GitHub.

@@

As an example, the template of the elementary space $T_1 = (1,1,1,1,1-x_1,x_2,x_1,0)$ represents all the rules which have at position 0 (always from right to left) state 0, at positions 4, 5, 6 and 7 the state 1, at positions 1 and 2 any state in the interval $ [0, k-1] $ and in position 3 the complementary state to the value of position 1. Therefore the template $T_1$ represents the set of elementary cellular automata in its decimal form $\{248,242,252,246\}$ or in binary form as presented below:
\begin{displaymath}
\{(1,1,1,1,1,0,0,0),(1,1,1,1,0,0,1,0),(1,1,1,1,1,1,0,0),(1,1,1,1,0,1,1,0)\}
\end{displaymath}

The process of finding all the rules represented by a template is called \textit{expansion}. There are implemented in the library \textit{CATemplates} \cite{CATemplates} several generators algorithms of templates representing rules with certain property. %TODO Inverter frase
 The template $T_{comp} = (1 - x_0, 1 - x_4, 1 - x_2, x_4, 1 - x_1, x_2, x_1, x_0)$ of the elementary CAs, for example, when expanded generates only rules that have a maximum internal symmetry by composition. But the template $T_{inv} = (1 - x_0, 1 - x_1, 1 - x_2, 1 - x_3, x_3, x_2, x_1, x_0)$ of the elementary CAs expands only the rules with color blind property.

Besides the possibility of expansion of templates, it has been developed by De Oliveira e Verardo \cite{deOliveira2014b} an algorithm to generate templates that represent the intersection between two sub-spaces of rules represented by templates. To illustrate, consider the search for a rule set that are both color blind and have maximum internal symmetry by composition. To find this rule set is necessary to perform two steps: first one must match the two templates that represent the desired properties, thus forming an equation system; then to solve the equations, relationships are generated between the variables that, when applied to templates received as input will result in the template representing the intersection.

@@

For example, we will continue the difference operation between $T_{inv}$ and $T_{con}$; to this end, consider the template $T_i = (1, 1 - x_1, 1 - x_2, 1 - x_1 - x_2, x_1 + x_2, x_2, x_1, 0)$ the intersection between them, for $k=2$.
The first step of the difference operation occurs normally and generates the templates: 
\begin{displaymath}
\{(x_7, x_6, x_5, x_4, x_3, x_2, x_1, 1),(x_7, x_6, x_5, x_1 + x_2, x_3, x_2, x_1, x_0)\}
\end{displaymath}
It is trivial to realize that any expansion of $T_i$ having the set of substitutions $\{x_1 = 1, x_2 = 1\}$ will cause the position $3$ and $4$ of the template display values that do not belong to the interval $[0,k-1]$.
Hence all templates that have $\{x_1 = 1, x_2 = 1\}$ are complementary to the template $T_i$. So it will generate the template $(x_7, x_6, x_5, x_4, x_3, 1, 1, x_0)$, which is the exception template of $T_i$.

Therefore every rule represented  by the template $T_{minuend}$ -- $T_{inv}$ in the given example -- and which is also represented by the exception template from the the intersection of $T_{minuend}$ with $T_{subtrahend}$ -- $T_i$ in this example - - should be represented by at least one of the templates resulting from the difference operation. For this, the algorithm that finds the difference between templates considers all exception templates found, intersects them with the $T_{minuend}$ and adds them to the set of templates obtained by the first steps of the difference operation. Thus, for the example presented above, the set of difference templates resultant is represented below:\begin{displaymath}
\begin{split}
\{(0, 1 - x_1, 1 - x_2, 1 - x_3, x_3, x_2, x_1, 1), \\
(1 - x_0, 1 - x_1, 1 - x_2, x_1 + x_2, 1 - x_1 - x_2, x_2, x_1, x_0), \\
(1 - x_0, 0, 0, 1 - x_3, x_3, 1, 1, x_0)\}
\label{eq:differenceR}
\end{split}
\end{displaymath}

@@

\section*{Acknowledgements}
\label{sec:agrdecimentos}
The authors thank Mackpesquisa, the Fapesp, and federal agencies CAPES and CNPq for different forms of support received during the development of this work.

% ---- Bibliography ----
\bibliography{bibliografia}
\bibliographystyle{splncs}

\end{document}
